
## Effective Data Wrangling

Import CEL data

```{r}
# Load libraries
library(affy)
library(limma)

# Read RAW Data
dir.path <- "E-GEOD-46602"
fns <- sort(list.celfiles(path=dir.path, full.names=TRUE))
data <- ReadAffy(filenames=fns, phenoData=NULL)
summary(data)

# Modify sample names
current_names <- sampleNames(data)
modified_names <- sub("_.*", "", current_names)
sampleNames(data) <- modified_names

sampleNames(data)

```

Import annotation

```{r}
# annotation
annotations <- read.table("E-GEOD-46602.sdrf.txt", header = TRUE, sep = "\t", row.names = 1)
rownames(annotations) <- gsub(" 1$", "", rownames(annotations))
ann <- annotations[, c(
  "Characteristics..age.",
  "Characteristics..organism.part."
)]

colnames(ann) <- c(
  "Age",
  "Organism_Part"
)
ann
```

Impute Missing Value

```{r}
colSums(is.na(ann))

# Impute missing values in Age columns with the median
num_cols <- c("Age")
for (col in num_cols) {
  ann[[col]][is.na(ann[[col]])] <- median(ann[[col]], na.rm = TRUE)
}

# Add FactorValue..margin_status. to ann
ann$Margin_Status <- annotations[rownames(ann), "FactorValue..margin_status."]

# Replace NA values with "Normal"
ann$Margin_Status[is.na(ann$Margin_Status)] <- "Normal"

ann$bcr <- annotations[rownames(ann), "Characteristics..bcr."]
ann$bcr[is.na(ann$bcr)] <- "Normal"

ann
```

Categorize class

```{r}
# Categorize samples
ann$Age_Group <- ifelse(ann$Age < 65, "Age<65", "Age≥65")
ann$Disease_Status <- ifelse(
  ann$Organism_Part == "benign prostate glands", 
  "Normal", 
  "Disease"
)

head(ann)

table(ann$Age_Group)
table(ann$Disease_Status)
table(ann$Margin_Status)
table(ann$bcr)
```

## Transformation and Normalization

RMA normalization

```{r}
# View a summary of the normalized data
dim(exprs(data))
```

```{r}
normalized_data <- rma(data)

dim(normalized_data)
```

```{r}
summary(exprs(normalized_data))
```

```{r}
dim(exprs(normalized_data))
```

```{r}
# Function to calculate M and A values
calculate_MA <- function(expr_matrix) {
  # Log2 transform the raw data
  expr_matrix <- log2(expr_matrix + 1)
  # Calculate M (log ratio) and A (average intensity)
  A <- rowMeans(expr_matrix)
  M <- expr_matrix[, 1] - rowMeans(expr_matrix[, -1])
  data.frame(A = A, M = M)
}

# Calculate M and A values for raw data
ma_raw <- calculate_MA(exprs(data))
```

Comparison between no normalization and rma normalization

```{r}
# Function to calculate M and A values
calculate_MA <- function(expr_matrix) {
  # Log2 transform if not already done
  if (any(expr_matrix > 100)) {
    expr_matrix <- log2(expr_matrix + 1)
  }
  # Calculate M (log ratio) and A (average intensity)
  A <- rowMeans(expr_matrix)
  M <- expr_matrix[, 1] - rowMeans(expr_matrix[, -1])
  data.frame(A = A, M = M)
}
# Generate MvA data for RMA normalized data
ma_rma <- calculate_MA(exprs(normalized_data))


library(ggplot2)
par(mfrow = c(1, 2))
ggplot(ma_raw, aes(x = A, y = M)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "MvA Plot - Before Normalization", x = "A", y = "M") +
  theme_minimal()

# Plot MvA for RMA normalized data
ggplot(ma_rma, aes(x = A, y = M)) +
  geom_point(alpha = 0.5, color = "black") +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "MvA Plot - RMA Normalization", x = "A", y = "M") +
  theme_minimal()

```

## Data noise filtering
```{r}
# Calculate the mean and standard deviation for each gene
gene_means <- rowMeans(exprs(normalized_data))
gene_sds <- apply(exprs(normalized_data), 1, sd)

# Calculate CV for each gene
gene_cvs <- gene_sds / gene_means

# Plot histogram of CVs
hist(gene_cvs, breaks = 50, col = "lightblue",
     main = "Distribution of Coefficient of Variation (CV)",
     xlab = "Coefficient of Variation")

```

```{r}
nrow(exprs(normalized_data))

cv_threshold <- quantile(gene_cvs, 0.75)

# Filter genes based on CV
filtered_data <- exprs(normalized_data)[gene_cvs > cv_threshold, ]

# Verify the dimensions after filtering
cat("Number of genes retained after CV filtering:", nrow(filtered_data), "\n")

```

```{r}
# Plot CVs of retained genes
filtered_cvs <- gene_cvs[gene_cvs > cv_threshold]

hist(filtered_cvs, breaks = 50, col = "lightgreen",
     main = "CV Distribution After Filtering",
     xlab = "Coefficient of Variation")


```

Retained the top 25% of transcripts based on their variance



## Exploratory analysis


```{r}
# Perform PCA on the filtered data
pca_res <- prcomp(t(filtered_data), scale. = TRUE)

# Plot PCA results
library(ggplot2)
pca_data <- as.data.frame(pca_res$x)
pca_data$Group <- ann[rownames(pca_res$x), ]$Disease_Status  

# Add labels to points
ggplot(pca_data, aes(x = PC1, y = PC2, color = Group, label = rownames(pca_data))) +
  geom_point(size = 2) +
  geom_text(vjust = -0.5, size = 3) +  # Add labels slightly above points
  labs(title = "PCA of Filtered Data", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()


```
```{r}
pca_data$Age <- ann[rownames(pca_res$x), ]$Age_Group  

# Add labels to points
ggplot(pca_data, aes(x = PC1, y = PC2, color = Age, label = rownames(pca_data))) +
  geom_point(size = 2) +
  geom_text(vjust = -0.5, size = 3) +  # Add labels slightly above points
  labs(title = "PCA of Filtered Data", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```
```{r}
pca_data$Margin <- annotations[rownames(pca_res$x), ]$FactorValue..margin_status.  

# Add labels to points
ggplot(pca_data, aes(x = PC1, y = PC2, color = Margin, label = rownames(pca_data))) +
  geom_point(size = 2) +
  geom_text(vjust = -0.5, size = 3) +  # Add labels slightly above points
  labs(title = "PCA of Filtered Data", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```

```{r}
library(pheatmap)

# Calculate correlation matrix
cor_matrix <- cor(filtered_data, use = "pairwise.complete.obs", method = "pearson")

# Subset and reorder ann to match the column names of cor_matrix
ann_aligned <- ann[colnames(cor_matrix), , drop = FALSE]

annotation_col <- data.frame(Group = ann_aligned$Disease_Status)
rownames(annotation_col) <- rownames(ann_aligned)
ordered_samples <- order(annotation_col$Group)

# Reorder cor_matrix and annotation_col
cor_matrix <- cor_matrix[ordered_samples, ordered_samples]
annotation_col <- annotation_col[ordered_samples, , drop = FALSE]

pheatmap(cor_matrix,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Sample Correlation Heatmap (Grouped by Disease Status)",
         annotation_col = annotation_col,
         cluster_cols = FALSE, 
         cluster_rows = FALSE,
         fontsize_row = 6,
         fontsize_col = 6) 
```


## Outlier assessment and removal
```{r}
# Calculate average correlation for each sample
avg_cor <- apply(cor_matrix, 1, mean)
# Plot average correlation
plot(c(1, length(avg_cor)), range(avg_cor), 
     type = "n", 
     xlab = "Sample", 
     ylab = "Average Correlation (r)", 
     main = "Average Correlation of Samples", 
     axes = FALSE)

# Add points to the plot
points(avg_cor, bg = "red", col = 1, pch = 21, cex = 1.25)

# Add X and Y axes
axis(1, at = 1:length(avg_cor), labels = colnames(cor_matrix), las = 2, cex.axis = 0.6)
axis(2)

# Add vertical lines for clarity
abline(v = seq(0.5, length(avg_cor) + 0.5, 1), col = "grey")

```

```{r}
# Calculate Mean, Standard Deviation, and CV for each sample
filtered_mean <- apply(log2(filtered_data), 2, mean)  # Mean expression per sample
filtered_sd <- sqrt(apply(log2(filtered_data), 2, var))  # Standard deviation per sample
filtered_cv <- filtered_sd / filtered_mean  # Coefficient of Variation (CV)

# Plot CV vs. Mean
plot(filtered_mean, filtered_cv, 
     main = "CV vs. Mean Plot for Filtered Data", 
     xlab = "Mean Expression (log2)", 
     ylab = "Coefficient of Variation (CV)", 
     col = 'blue', cex = 1.5, type = "n")  # Create an empty plot

# Add points to the plot
points(filtered_mean, filtered_cv, bg = "lightblue", col = 1, pch = 21)

# Label points with sample IDs
text(filtered_mean, filtered_cv, labels = colnames(filtered_data), cex = 0.7)

```
## Clustering 
```{r}
# Define outliers
outliers <- c("GSM1133140", "GSM1133169")

# Remove outliers from the dataset
dat <- filtered_data[, !colnames(filtered_data) %in% outliers]

# Verify dimensions
dim(dat)

```
```{r}
# Calculate distance matrix
distance_matrix <- dist(t(dat), method = "euclidean")  # Transpose for sample-wise distance

# Perform hierarchical clustering
hc <- hclust(distance_matrix, method = "complete")


library(dendextend)

ann <- ann[match(colnames(dat), rownames(ann)), ]
# Assign colors based on disease status and age group
disease_colors <- ifelse(ann$Disease_Status == "Disease", "red", "blue")  # Disease: red, Normal: blue
age_colors <- ifelse(ann$Age_Group == "Age<65", "green", "purple")       # Age<65: green, Age≥65: purple
names(disease_colors) <- hc$labels
disease_colors
```

HCA
```{r}
library(dendextend)

dend_disease <- as.dendrogram(hc)

# Reorder disease_colors based on the labels in the dendrogram
disease_colors_ordered <- disease_colors[order.dendrogram(dend_disease)]

dend_disease <- dend_disease %>%
  set("labels_colors", value = disease_colors_ordered) %>%
  set("labels_cex", 0.6)  # Reduce label font size

# Plot the dendrogram
plot(dend_disease, main = "Clustering by Disease Status", xlab = "Samples", sub = "")

# Add a legend
legend("topright", legend = c("Disease", "Normal"), fill = c("red", "blue"), 
       title = "Disease Status", bty = "n", cex = 0.8)


```
```{r}
margin_colors <- ifelse(
  annotations$FactorValue..margin_status. == "Positive", "red", 
  ifelse(annotations$FactorValue..margin_status. == "Negative", "orange", NA)
)

margin_colors <- margin_colors[rownames(annotations) %in% colnames(dat)]
names(margin_colors) <- rownames(annotations)[rownames(annotations) %in% colnames(dat)]

dend_margin <- as.dendrogram(hc)
labels_order <- labels(dend_margin) 
# Filter and reorder margin_colors to match dendrogram labels
margin_colors_aligned <- margin_colors[labels_order]
names(margin_colors_aligned) <- labels_order 
dend_margin <- dend_margin %>%
  set("labels_colors", value = margin_colors_aligned) %>%
  set("labels_cex", 0.6)  # Adjust label font size

# Plot the dendrogram
plot(dend_margin, main = "Clustering by Margin Status", xlab = "Samples", sub = "")

# Add a legend
legend("topright", legend = c("Positive", "Negative","Normal"), fill = c("red", "orange","black"), 
       title = "Margin Status", bty = "n", cex = 0.8)
```
```{r}
bcr_colors <- ifelse(
  ann$bcr == "YES", "red", 
  ifelse(ann$bcr == "NO", "orange", NA)
)

bcr_colors <- bcr_colors[rownames(ann) %in% colnames(dat)]
names(bcr_colors) <- rownames(ann)[rownames(ann) %in% colnames(dat)]

dend_bcr <- as.dendrogram(hc)
labels_order <- labels(dend_bcr) 
# Filter and reorder margin_colors to match dendrogram labels
bcr_colors_aligned <- bcr_colors[labels_order]
names(bcr_colors_aligned) <- labels_order 
dend_bcr <- dend_bcr %>%
  set("labels_colors", value = bcr_colors_aligned) %>%
  set("labels_cex", 0.6)  # Adjust label font size

# Plot the dendrogram
plot(dend_bcr, main = "Clustering by Biochemical Recurrence", xlab = "Samples", sub = "")

# Add a legend
legend("topright", legend = c("YES", "NO","Normal"), fill = c("red", "orange","black"), 
       title = "Biochemical Recurrence", bty = "n", cex = 0.8)
```


K-Means
```{r}
pca.res <- prcomp(t(dat), scale. = TRUE)
# Use the first 2 principal components for clustering
pca.data <- pca.res$x[, 1:2]  # Select PC1 and PC2
pca.data <- as.data.frame(pca.data)

# Perform k-means clustering
set.seed(123)  # Ensure reproducibility
kmeans_res <- kmeans(pca.data, centers = 3, iter.max = 10)  

# Add cluster assignments to PCA data
pca.data$Cluster <- as.factor(kmeans_res$cluster)
# Prepare cluster centers
cluster_centers <- as.data.frame(kmeans_res$centers)
colnames(cluster_centers) <- c("PC1", "PC2") 

library(ggrepel)

pca.data$Disease_Status  <- ann[rownames(pca.res$x), ]$Disease_Status
pca.data$Age_Group <- ann[rownames(pca.res$x), ]$Age_Group

library(ggplot2)

# Plot k-means clusters with Disease Status and Age Group
ggplot(pca.data, aes(x = PC1, y = PC2, color = Cluster, shape = Age_Group, label = Disease_Status)) +
  geom_point(size = 3) +  # Plot data points with shapes for Age Groups
  geom_point(data = cluster_centers, aes(x = PC1, y = PC2), 
             inherit.aes = FALSE, color = "black", shape = 8, size = 4) +  # Add cluster centers
  geom_text_repel(size = 3, max.overlaps = 15) +  # Add Disease Status labels
  labs(title = "K-Means Clustering - age", 
       x = "Principal Component 1", 
       y = "Principal Component 2", 
       color = "Cluster", 
       shape = "Age Group") +
  theme_minimal()
```

```{r}
pca.data$Margin_Status <- ifelse(
  pca.data$Disease_Status == "Disease",  # Only include for Disease samples
  annotations$FactorValue..margin_status.[match(rownames(pca.data), rownames(annotations))],
  NA
)

library(ggplot2)
library(ggrepel)

# Plot k-means clusters with labels only for samples with Margin_Status
ggplot(pca.data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, aes(shape = Disease_Status)) +  # Shape differentiates Disease/Normal
  geom_point(data = cluster_centers, aes(x = PC1, y = PC2), 
             inherit.aes = FALSE, color = "black", shape = 8, size = 4) +  # Add cluster centers
  geom_text_repel(aes(label = ifelse(!is.na(Margin_Status), Margin_Status, "")), 
                  size = 3, max.overlaps = 15, na.rm = TRUE) +  # Label only if Margin_Status is not NA
  labs(title = "K-Means Clustering (k = 3) with Margin Status Labels", 
       x = "Principal Component 1", 
       y = "Principal Component 2", 
       color = "Cluster", 
       shape = "Disease Status") +
  theme_minimal()

```

```{r}
# Add BRC (Biochemical Recurrence) Status to PCA Data
pca.data$bcr <- ifelse(
  pca.data$Disease_Status == "Disease",  # Include only for Disease samples
  annotations$Characteristics..bcr.[match(rownames(pca.data), rownames(annotations))],
  NA
)

library(ggplot2)
library(ggrepel)

# Plot k-means clusters with BRC labels
ggplot(pca.data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, aes(shape = Disease_Status)) +  # Shape differentiates Disease/Normal
  geom_point(data = cluster_centers, aes(x = PC1, y = PC2), 
             inherit.aes = FALSE, color = "black", shape = 8, size = 4) +  # Add cluster centers
  geom_text_repel(aes(label = ifelse(!is.na(bcr), bcr, "")), 
                  size = 3, max.overlaps = 15, na.rm = TRUE) +  # Label only if BRC is not NA
  labs(title = "K-Means Clustering (k = 3) with BCR Labels", 
       x = "Principal Component 1", 
       y = "Principal Component 2", 
       color = "Cluster", 
       shape = "Disease Status") +
  theme_minimal()

```

## Normality and variance assessment and differential test selection defended
```{r}
# Load necessary libraries
library(stats)

# Perform Shapiro-Wilk test for normality
set.seed(1129)
genes_to_test <- sample(rownames(dat), 100)  # Test 100 random genes
shapiro_results <- apply(dat[genes_to_test, ], 1, function(x) {
  shapiro.test(x)$p.value
})

mean(shapiro_results > 0.05)

normality_pvalues <- apply(dat, 1, function(x) {
  if (length(unique(x)) > 3) { # Shapiro-Wilk requires at least 4 unique values
    shapiro.test(x)$p.value
  } else {
    NA
  }
})
cat("Number of genes with normal distribution:", sum(normality_pvalues > 0.05, na.rm = TRUE), "\n")
```



Variance Analysis
```{r}
library(car)
group_labels <- factor(ann[colnames(dat), "Disease_Status"], levels = c("Normal", "Disease"))

# Perform Levene's test for each gene
levene_results <- apply(dat, 1, function(x) {
  if (length(unique(group_labels)) > 1) {  # Ensure multiple groups exist
    leveneTest(x ~ group_labels)$`Pr(>F)`[1]  # Extract p-value
  } else {
    NA
  }
})

# Summary of results
cat("Number of genes with equal variance (p >= 0.05):", sum(levene_results >= 0.05, na.rm = TRUE), "\n")
cat("Number of genes with unequal variance (p < 0.05):", sum(levene_results < 0.05, na.rm = TRUE), "\n")
```

## Differential testing with test selected for use

age
```{r}
# Ensure Age_Group is numeric
age_numeric <- as.numeric(ann$Age)
spearman_results <- apply(dat, 1, function(x) {
  cor(x, age_numeric, method = "spearman")
})

# Perform Spearman correlation test for each gene
spearman_pvalues <- apply(dat, 1, function(x) {
  cor.test(x, age_numeric, method = "spearman", exact = FALSE)$p.value  # Use asymptotic approximation
})

# Adjust p-values using FDR
spearman_pvalues_adjusted <- p.adjust(spearman_pvalues, method = "fdr")

# Extract significant genes (FDR < 0.05)
significant_age_genes <- names(spearman_pvalues_adjusted[spearman_pvalues_adjusted < 0.05])

# Output the number and names of significant genes
cat("Number of significant genes (Spearman, Age):", length(significant_age_genes), "\n")
```



Wilcoxon Rank-Sum Test

```{r}
# Perform Wilcoxon rank-sum test with continuity correction (asymptotic approximation)
disease_vs_normal_pvalues <- apply(dat, 1, function(x) {
  wilcox.test(x ~ group_labels, exact = FALSE)$p.value  # Use approximation for tied data
})


significant_genes <- names(disease_vs_normal_pvalues[disease_vs_normal_pvalues < 0.05])
cat("Number of significant genes (Disease vs. Normal):", length(significant_genes), "\n")

```
```{r}
# Define group labels for Margin Status and BCR
margin_status <- factor(ann[colnames(dat), "Margin_Status"])

# Perform Kruskal-Wallis Test for Margin Status
kw_pvalues_margin <- apply(dat, 1, function(x) {
  kruskal.test(x ~ margin_status)$p.value
})

# Count significant genes
significant_genes_margin <- sum(kw_pvalues_margin < 0.05)
cat("Number of significant genes (Kruskal-Wallis, Margin Status):", significant_genes_margin, "\n")

```

```{r}
bcr_status <- factor(ann[colnames(dat), "bcr"])

# Perform Kruskal-Wallis Test for BCR
kw_pvalues_bcr <- apply(dat, 1, function(x) {
  kruskal.test(x ~ bcr_status)$p.value
})

# Count significant genes
significant_genes_bcr <- sum(kw_pvalues_bcr < 0.05)
cat("Number of significant genes (Kruskal-Wallis, BCR):", significant_genes_bcr, "\n")

```

```{r}
# Histogram for Disease vs. Normal (Wilcoxon Test)
hist(disease_vs_normal_pvalues, breaks = 50, col = "lightblue",
     main = "Wilcoxon Test Histogram of P-Values (Normal vs Disease)",
     xlab = "P-Values",
     ylab = "Frequency")

# Histogram for Margin Status (Kruskal-Wallis Test)
hist(kw_pvalues_margin, breaks = 50, col = "lightcoral",
     main = "Kruskal-Wallis Test Histogram of P-Values (Margin Status)",
     xlab = "P-Values",
     ylab = "Frequency")

# Histogram for BCR (Kruskal-Wallis Test)
hist(kw_pvalues_bcr, breaks = 50, col = "lightgreen",
     main = "Kruskal-Wallis Test Histogram of P-Values (BCR)",
     xlab = "P-Values",
     ylab = "Frequency")

```
## Multiple comparison correction

```{r}
if (!require("multtest")) install.packages("multtest")
library(multtest)

group_labels <- factor(ann[colnames(dat), "Disease_Status"], levels = c("Normal", "Disease"))

# Perform permutation-based correction using the MinP method
resP <- mt.minP(dat, group_labels, test = "t", side = "abs")$rawp

# Adjust p-values using multiple methods
adjusted_pvalues <- mt.rawp2adjp(resP, proc = c("Bonferroni", "Holm", "BH", "BY", "Hochberg"))

# Extract adjusted p-values and sort by raw p-values
adjusted_pvalues_sorted <- adjusted_pvalues$adjp[order(adjusted_pvalues$index), ]
```

```{r}
plot(
  1:length(resP),
  sort(resP),  # Raw p-values
  type = "l",
  col = "black",
  lwd = 2,
  xlab = "Number of Rejected Hypotheses",
  ylab = "Adjusted p-values",
  main = "Performance Comparison of MCC Methods"
)
lines(
  1:length(resP),
  sort(adjusted_pvalues_sorted[, "BH"]),
  col = "blue",
  lwd = 2
)
lines(
  1:length(resP),
  sort(adjusted_pvalues_sorted[, "Bonferroni"]),
  col = "red",
  lwd = 2
)
lines(
  1:length(resP),
  sort(adjusted_pvalues_sorted[, "Holm"]),
  col = "green",
  lwd = 2
)
lines(
  1:length(resP),
  sort(adjusted_pvalues_sorted[, "BY"]),
  col = "purple",
  lwd = 2
)
lines(
  1:length(resP),
  sort(adjusted_pvalues_sorted[, "Hochberg"]),
  col = "orange",
  lwd = 2
)

# Add legend
legend(
  "bottomright",
  legend = c("Raw p-values", "BH", "Bonferroni", "Holm", "BY", "Hochberg"),
  col = c("black", "blue", "red", "green", "purple", "orange"),
  lwd = 2
)
```

```{r}
# Extract adjusted p-values for each method
raw_pvalues <- adjusted_pvalues_sorted[, "rawp"]
bonferroni_pvalues <- adjusted_pvalues_sorted[, "Bonferroni"]
holm_pvalues <- adjusted_pvalues_sorted[, "Holm"]
bh_pvalues <- adjusted_pvalues_sorted[, "BH"]
by_pvalues <- adjusted_pvalues_sorted[, "BY"]
hochberg_pvalues <- adjusted_pvalues_sorted[, "Hochberg"]

# Calculate number of significant genes for each method
significant_counts <- data.frame(
  Method = c("Raw", "Bonferroni", "Holm", "BH", "BY", "Hochberg"),
  Significant_Genes = c(
    sum(raw_pvalues < 0.05),
    sum(bonferroni_pvalues < 0.05),
    sum(holm_pvalues < 0.05),
    sum(bh_pvalues < 0.05),
    sum(by_pvalues < 0.05),
    sum(hochberg_pvalues < 0.05)
  )
)

# Display the counts
print(significant_counts)

```

Choose BH

```{r}
# Histogram of BH-adjusted p-values
significant_bh_pvalues <- adjusted_pvalues_sorted[, "BH"]
hist(
  significant_bh_pvalues,
  breaks = 50,
  col = "lightblue",
  main = "Distribution of Adjusted P-Values (BH)",
  xlab = "Adjusted P-Value",
  ylab = "Frequency"
)

```

```{r}
# Adjust p-values using BH for Disease vs. Normal
bh_adjusted_pvalues_disease <- p.adjust(disease_vs_normal_pvalues, method = "BH")

# Select significant genes (FDR < 0.05)
significant_genes_bh_disease <- names(bh_adjusted_pvalues_disease[bh_adjusted_pvalues_disease < 0.05])

# Calculate threshold
threshold_bh_disease <- max(bh_adjusted_pvalues_disease[bh_adjusted_pvalues_disease < 0.05])

# Output
cat("Number of significant genes (BH, Disease vs. Normal):", length(significant_genes_bh_disease), "\n")
cat("Threshold value (BH, Disease vs. Normal):", threshold_bh_disease, "\n")

```
```{r}
# Adjust p-values using BH for Margin Status
bh_adjusted_pvalues_margin <- p.adjust(kw_pvalues_margin, method = "BH")

# Select significant genes (FDR < 0.05)
significant_genes_bh_margin <- names(bh_adjusted_pvalues_margin[bh_adjusted_pvalues_margin < 0.05])

# Calculate threshold
threshold_bh_margin <- max(bh_adjusted_pvalues_margin[bh_adjusted_pvalues_margin < 0.05])

# Output
cat("Number of significant genes (BH, Margin Status):", length(significant_genes_bh_margin), "\n")
cat("Threshold value (BH, Margin Status):", threshold_bh_margin, "\n")

```


```{r}
# Adjust p-values using BH for BCR
bh_adjusted_pvalues_bcr <- p.adjust(kw_pvalues_bcr, method = "BH")

# Select significant genes (FDR < 0.05)
significant_genes_bh_bcr <- names(bh_adjusted_pvalues_bcr[bh_adjusted_pvalues_bcr < 0.05])

# Calculate threshold
threshold_bh_bcr <- max(bh_adjusted_pvalues_bcr[bh_adjusted_pvalues_bcr < 0.05])

# Output
cat("Number of significant genes (BH, BCR):", length(significant_genes_bh_bcr), "\n")
cat("Threshold value (BH, BCR):", threshold_bh_bcr, "\n")

```

## Confirmatory analysis 
```{r}
shared_significant_genes <- Reduce(intersect, list(
    significant_genes_bh_disease,
    significant_genes_bh_margin,
    significant_genes_bh_bcr
))

subset_data <- dat[shared_significant_genes, ]

# Display the dimensions of the subset data
cat("Subset data dimensions (Shared Significant Genes):", dim(subset_data), "\n")
```

```{r}
pca_res <- prcomp(t(subset_data), scale. = TRUE)
pca_data <- as.data.frame(pca_res$x)
pca_data$Group <- ann[colnames(subset_data), "Disease_Status"]

# Visualize PCA
library(ggplot2)
ggplot(pca_data, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 3) +
  labs(title = "PCA of Significant Genes", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```


```{r}
pca_summary <- summary(pca_res)
pca_variance <- pca_summary$importance[2,]

# scree plot
plot(pca_variance, type = "b", pch = 19, col = "blue",
     main = "Scree Plot of PCA",
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained")

variance_first_two <- sum(pca_variance[1:2]) * 100
cat("The first two principal components explain approximately", round(variance_first_two, 2), "% of the variability in the data.\n")
```


```{r}
# Correlation matrix
cor_matrix <- cor(subset_data, method = "pearson")

annotation_col.sub <- data.frame(Group = ann[colnames(subset_data), "Disease_Status"])
rownames(annotation_col.sub) <- colnames(subset_data)  # Ensure matching row names

library(pheatmap)
pheatmap(
  cor_matrix,
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = "Sample Correlation Heatmap",
  annotation_col = annotation_col.sub,
  fontsize_row = 6,
  fontsize_col = 6
)


```


## Classification Modeling

```{r}
library(caret)  # For splitting data and running models

# Split data into training and testing sets
set.seed(123)
trainIndex <- createDataPartition(ann[colnames(subset_data), "Disease_Status"], p = 0.75, list = FALSE)
train_data <- subset_data[, trainIndex]
test_data <- subset_data[, -trainIndex]

# Extract labels
train_labels <- ann[colnames(train_data), "Disease_Status"]
test_labels <- ann[colnames(test_data), "Disease_Status"]

margin_labels <- factor(ann[colnames(subset_data), "Margin_Status"])
bcr_labels <- factor(ann[colnames(subset_data), "bcr"])

train_margin <- margin_labels[trainIndex]
test_margin <- margin_labels[-trainIndex]

train_bcr <- bcr_labels[trainIndex]
test_bcr <- bcr_labels[-trainIndex]
```

SVM
```{r}
library(e1071)

# Train SVM model
svm_model <- svm(t(train_data), as.factor(train_labels), kernel = "linear", scale = TRUE)

# Predict on test data
svm_predictions <- predict(svm_model, t(test_data))

# Evaluate SVM performance
svm_conf_matrix <- confusionMatrix(as.factor(svm_predictions), as.factor(test_labels))
print(svm_conf_matrix)

```
```{r}
svm_model_margin <- svm(t(train_data), as.factor(train_margin), kernel = "linear", scale = TRUE)
svm_predictions_margin <- predict(svm_model_margin, t(test_data))
svm_conf_matrix_margin <- confusionMatrix(as.factor(svm_predictions_margin), as.factor(test_margin))
print(svm_conf_matrix_margin)


```

```{r}
# Train SVM for BCR
svm_model_bcr <- svm(t(train_data), as.factor(train_bcr), kernel = "linear", scale = TRUE)
svm_predictions_bcr <- predict(svm_model_bcr, t(test_data))
svm_conf_matrix_bcr <- confusionMatrix(as.factor(svm_predictions_bcr), as.factor(test_bcr))
print(svm_conf_matrix_bcr)
```




```{r}
# Perform PCA on test data
test_pca <- prcomp(t(test_data), scale. = TRUE)
pca_test_data <- as.data.frame(test_pca$x[, 1:2])  # Use PC1 and PC2
pca_test_data$Actual_Class <- test_labels
pca_test_data$Predicted_Class <- svm_predictions

library(ggplot2)

# PCA scatter plot with predicted classes
ggplot(pca_test_data, aes(x = PC1, y = PC2, color = Predicted_Class, shape = Actual_Class)) +
  geom_point(size = 3) +
  labs(
    title = "SVM Classification Results (PCA Projection)",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("red", "blue"), name = "Predicted Class") +
  scale_shape_manual(values = c(16, 17), name = "Actual Class")


```

```{r}
pca_test_data$Actual_Margin <- test_margin
pca_test_data$Predicted_Margin <- svm_predictions_margin

ggplot(pca_test_data, aes(x = PC1, y = PC2, color = Predicted_Margin, shape = Actual_Margin)) +
  geom_point(size = 3) +
  labs(
    title = "SVM Classification for Margin Status (PCA Projection)",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("green", "orange", "purple"), name = "Predicted Margin Status") +
  scale_shape_manual(values = c(16, 17, 18), name = "Actual Margin Status")
```

```{r}
pca_test_data$Actual_BCR <- test_bcr
pca_test_data$Predicted_BCR <- svm_predictions_bcr
ggplot(pca_test_data, aes(x = PC1, y = PC2, color = Predicted_BCR, shape = Actual_BCR)) +
  geom_point(size = 3) +
  labs(title = "SVM Classification for BCR",
       x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()
```
```{r}
disease_labels <- factor(ann[colnames(dat), "Disease_Status"], levels = c("Normal", "Disease"))
# Calculate the mean expression for each gene in Disease and Normal samples
mean_disease <- rowMeans(dat[, disease_labels == "Disease"])
mean_normal <- rowMeans(dat[, disease_labels == "Normal"])

log_fold_change <- log2(mean_disease + 1) - log2(mean_normal + 1)

sorted_genes <- sort(log_fold_change, decreasing = TRUE)

# Top 5 genes with highest log fold change (positive)
top_positive_genes <- head(sorted_genes, 5)
cat("Top 5 genes with the highest positive log fold change:\n")
print(top_positive_genes)

# Top 5 genes with lowest log fold change (negative)
top_negative_genes <- tail(sorted_genes, 5)
cat("Top 5 genes with the lowest log fold change:\n")
print(top_negative_genes)

```


```{r}

library(hgu133plus2.db)

# Define probe IDs
probe_ids <- c("242138_at", "232572_at", "1563392_at", "205422_s_at", "232575_at")

# Map probe IDs to gene symbols
gene_symbols <- mapIds(hgu133plus2.db, keys = probe_ids, column = "SYMBOL", keytype = "PROBEID")
print(gene_symbols)

# Map probe IDs to Entrez IDs
entrez_ids <- mapIds(hgu133plus2.db, keys = probe_ids, column = "ENTREZID", keytype = "PROBEID")
print(entrez_ids)

```

```{r}
library(hgu133plus2.db)

# Define probe IDs
probe_ids <- c("205765_at", "202504_at", "226755_at", "216158_at", "1552511_a_at")

# Map probe IDs to gene symbols
gene_symbols <- mapIds(hgu133plus2.db, keys = probe_ids, column = "SYMBOL", keytype = "PROBEID")
print(gene_symbols)

# Map probe IDs to Entrez IDs
entrez_ids <- mapIds(hgu133plus2.db, keys = probe_ids, column = "ENTREZID", keytype = "PROBEID")
print(entrez_ids)
```

